---
title: "Reproducible Research: Peer Assessment 1"
author: Haarkin898
date :  October 13 2020
output:
      html_document:
        keep_md: true
        self_contained: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries needed for the analysis (Quest 1)
```{r load libraries}
       
        library(dplyr)
        library(lubridate)
        library(ggplot2)
```

## read in the data set and assign it to an Object
```{r loan and explore dataset}
        ACM = read.csv("./activity.csv")
        head(ACM, 3)
        str(ACM)
```        

## refine the date variable, convert to date class and add as variable
```{r ref}       
        A <- as.Date(ACM$date, format = "%Y-%m-%d")
        class(A)
        ACM$A <- A
        ACM = rename(ACM, Date = A)
        ACM$date = NULL
```

## Group data by Date and analyze 
```{r Steps per Day}
        ACMbyDate = group_by(ACM, Date)
        ACMsum = summarise(ACMbyDate , steps = sum(steps, na.rm = TRUE) )
```       

## Plot histogram of above analysis
```{r HistogramACM}
        with(ACMsum, hist(steps, main = "Total number of steps per day", 
                          xlab ="Steps per day"))
```       

## Determine mean and median steps per day 
```{r summaryACM}
 ACMsum = rename(ACMsum, TotalSteps = steps)
 Mean = mean(ACMsum$TotalSteps)
 Median = median(ACMsum$TotalSteps)
 
``` 

The mean and median of the total number of steps taken per day are `r Mean` and `r Median` respectively

## Average daily activity time series  

```{r Int}       
        ACMbyInterval = group_by(ACM, interval)
        ACMbyInterval = summarise(ACMbyInterval , steps = mean(steps, 
                                                               na.rm = TRUE) )
```      
        
## Plot analysis from above
```{r plot}
        with(ACMbyInterval, plot(interval, steps, type = "l", 
        main = "Average Daily Activity Pattern", xlab = "Time Interval"
                                 , ylab = "Average Steps per Interval"))
```
        
## interval with the maximum number of steps 
```{r maxInt }
        maxInterval = ACMbyInterval[which.max(ACMbyInterval$steps),]$interval
        
```

The interval with the maximum number of steps is `r maxInterval`

## In-filling missing values with interval mean 
filling with daily mean is unlikely to change the output of the analysis as this will only return the same number of NA as the original dataset.
The dplyr library was used in this analysis
```{r replaceNA }       
        sumACM = sum(is.na(ACM))
        
        impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
        
        ACM3 <- ACM
        
        ACM3 = ACM3 %>%
                group_by(interval) %>%
                mutate(
                        steps = impute.mean(steps)
                )
        
        sumACM3 = sum(is.na(ACM3))
        
```

The sum of NAs in the original data, ACM, and the infilled data, ACM3, are 
`r sumACM` and `r sumACM3` respectively. 

## Histogram Plot of refined dataset ACM3 with respect to Total number of Steps per day

Group data by Date and analyze 
```{r ACM3}
        ACM3byDate = group_by(ACM3, Date)
        ACM3sum = summarise(ACM3byDate , steps = sum(steps, na.rm = TRUE) )
        Mean2 = mean(ACM3sum$steps)
        Median2 = median(ACM3sum$steps)
```
The mean and median of the total number of steps taken per day are `r Mean2` and 
`r Median2` respectively. This is slightly different from `r Mean` and `r Median` 
from the prior analysis. In this case the mean and the median are now the same 
and higher in value. This process has change the skewness from left skewed to
a symmetrical dataset.

## Plot histogram of above analysis 
```{r HistogramACM3 }
        with(ACM3sum, hist(steps, main = "Total number of steps per day", 
                          xlab ="Steps per day"))
        
```
        
## Time series analysis based on day type i.e. weekday vs weekends
        
```{r Daytype}
        Day <- weekdays(ACM3$Date)
        ACM3$Day = Day
        ACM3$Daytype <- ifelse(ACM3$Day %in% c("Saturday", "Sunday"), "weekend",
                               "weekday")
        ACM3$Daytype = factor(ACM3$Daytype)
        
        ACMbyInt = group_by(ACM3, interval, Daytype)
        ACMbyInt = summarize(ACMbyInt, steps = mean(steps))
        
```
##Plot
```{r ggplot}
        g <- ggplot(ACMbyInt, aes(x=interval, y=steps)) +
                geom_line() + facet_grid(Daytype~.) +
                labs (x = "Time Interval", y = "Average Steps per Interval") +
                labs (title = "Average Daily Activity Pattern by Day Type")
        g
```
